---
title: 'Bios 6301: Assignment 2'
author: "Elizabeth Sigworth"
date: "due October 10, 2016"
output: pdf_document
---

*(informally) Due Tuesday, 20 September, 1:00 PM*

50 points total.

This assignment won't be submitted until we've covered Rmarkdown.
Create R chunks for each question and insert your R code appropriately.
Check your output by using the `Knit PDF` button in RStudio.

#####1. **Working with data** In the `datasets` folder on the course GitHub repo, you will find a file called `cancer.csv`, which is a dataset in comma-separated values (csv) format. This is a large cancer incidence dataset that summarizes the incidence of different cancers for various subgroups. (18 points)

    1. Load the data set into R and make it a data frame called `cancer.df`. (2 points)
    
```{r}
cancer.df <- read.csv("/var/folders/kp/zlsf12h14y92__lp668ljv2m0000gn/T//RtmpKa1QwJ/data2c844f081265", stringsAsFactors=FALSE)
```

    2. Determine the number of rows and columns in the data frame. (2)
```{r}
nrow(cancer.df) # number of rows
ncol(cancer.df) # number of columns
```

    3. Extract the names of the columns in `cancer.df`. (2)
```{r}
names(cancer.df)
```

    4. Report the value of the 3000th row in column 6. (2)
```{r}
cancer.df[3000,6]
```

    5. Report the contents of the 172nd row. (2)
```{r}
cancer.df[172,]
```

    6. Create a new column that is the incidence *rate* (per 100,000) for each row.(3)
```{r}
cancer.df$inc.rate <- cancer.df$incidence/(cancer.df$population/100000)
```

    7. How many subgroups (rows) have a zero incidence rate? (2)
```{r}
length(which(cancer.df$inc.rate == 0))
```

    8. Find the subgroup with the highest incidence rate.(3)
```{r}
cancer.df[which(cancer.df$inc.rate == max(cancer.df$inc.rate)),]
```

#####2. **Data types** (10 points)

1. Create the following vector: `x <- c("5","12","7")`. Which of the following commands will produce an error message? For each command, Either explain why they should be errors, or explain the non-erroneous result. (4 points)
```{r}
x <- c("5","12","7")
```
            
max(x): This will return "7", since character vectors are sorted based on the first character in the string, so "7" is read as greater than the "1" in the start of "12"

sort(x): This returns the vector in order "12" "5" "7", again sorting on the first character in the string, which would be "1" and then "5" and then "7"

sum(x): This command returns an error, since the sum command cannot be run on type "character".


2. For the next two commands, either explain their results, or why they should produce errors. (3 points)

y <- c("5",7,12): This command will create a vector of character type, since R puts all elements of vectors into the same type, so in this case converts 7 and 12 to "7" and "12".

y[2] + y[3]: Because vector y was stored as a character type, attempting to sum these two elements of y will result in an error, since characters cannot be summed. 


3. For the next two commands, either explain their results, or why they should produce errors. (3 points)

z <- data.frame(z1="5",z2=7,z3=12): This creates a data frame with one row and three columns; column z1 contains the character "5", column z2 contains the integer 7, and column z3 contains the integer 12.

z[1,2] + z[1,3]: This adds the elements in row 1 column 2 and row 1 column 3, which are the integers 7 and 12, giving a result of 19. 


#####3. **Data structures** Give R expressions that return the following matrices and vectors (*i.e.* do not construct them manually). (3 points each, 12 total)

1. $(1,2,3,4,5,6,7,8,7,6,5,4,3,2,1)$
```{r}
c(seq(1:8),rev(seq(7:1)))
```
    
2. $(1,2,2,3,3,3,4,4,4,4,5,5,5,5,5)$
    ```{r}
    rep(1:5,1:5)
    ```

3. $\begin{pmatrix}
      0 & 1 & 1  \\
      1 & 0 & 1  \\
      1 & 1 & 0  \\
    \end{pmatrix}$

```{r}
mat <- matrix(rep(1,9),nrow=3)
for (i in 1:3) {
  for (j in 1:3) {
    if (i==j) {
      mat[i,j] <- 0
    }
  }
}
mat
```

4. $\begin{pmatrix}
      1 & 2 & 3 & 4 \\
      1 & 4 & 9 & 16 \\
      1 & 8 & 27 & 64  \\
      1 & 16 & 81 & 256 \\
      1 & 32 & 243 & 1024  \\
    \end{pmatrix}$
    
```{r}
x <- 1:4
matrix(c(x,x^2,x^3,x^4,x^5),nrow=5,ncol=4,byrow=TRUE)
```


4. **Basic programming** (10 points)

1. Let $h(x,n)=1+x+x^2+\ldots+x^n = \sum_{i=0}^n x^i$. Write an R program to calculate $h(x,n)$ using a `for` loop. (5 points)

```{r}
h.x.n <- function(x,n) {
  h <- 0
  for (i in 0:n) {
    h <- h + x^n
  }
  return(h)
}
```

2. If we list all the natural numbers below 10 that are multiples of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23. Write an R program to perform the following calculations. (5 points)
```{r}
#let a and b be the multiples of interest, and x be the value below which we are interested
sum.mult <- function(a,b,x) {
  total <- 0
  for (i in 1:x-1) {
    if (i%%a==0 | i%%b==0) {
      total <- total + i
    }
  }
  return(total)
}
```

(a) Find the sum of all the multiples of 3 or 5 below 1,000. (3, [euler1])

```{r}
sum.mult(3,5,1000)
```

(b) Find the sum of all the multiples of 4 or 7 below 1,000,000. (2)

```{r}
sum.mult(4,7,1000000)
```

(c) Each new term in the Fibonacci sequence is generated by adding the previous two terms. By starting with 1 and 2, the first 10 terms will be $(1, 2, 3, 5, 8, 13, 21, 34, 55, 89)$. Write an R program to calculate the sum of the first 15 even-valued terms. (5 bonus points, [euler2])

```{r}

## Generate the first 50 terms of the Fibonacci sequence
fibbo <- NULL
fibbo[1] <- 1
fibbo[2] <- 2
for (i in 3:50) {
  fibbo[i] <- fibbo[i-1] + fibbo[i-2]
}
fibbo

## Select only even terms from the first 50
fibbo.even <- fibbo[which(fibbo%%2==0)]
fibbo.even
length(fibbo.even)

## Since 17 of the first 50 terms were even, we select the first 15 and find the sume of these terms
first.15 <- sum(fibbo.even[1:15])
first.15
```


Some problems taken or inspired by projecteuler.

[euler1]:https://projecteuler.net/problem=1
[euler2]: https://projecteuler.net/problem=2